% Setting Up R
% Jared Knowles

# Materials
- [Download R](http://cran.r-project.org/)
- [RStudio](http://www.rstudio.org/) 
- We'll also want to install a few basic packages. This can be done on the fly later, but a few include `ggplot2`;`knitr`;`plyr`

# Step 1: Install R
- Simply run the installer. If you have a 64-bit machine, choose 64 bit R, if you don't know, just choose the default
- The install should be less than 100mb
- There is a short video on the next slide for help installing on Windows machines


# Installing R on Windows
<p align='center'><iframe src="http://www.screenr.com/embed/kzT8" width="650" height="396" frameborder="0"></iframe></p>


# Step 2: RStudio
- Now run the RStudio installer
- RStudio will automatically find your R installation

# Step 3: Install packages
First, watch this tutorial<br>

<iframe src="http://www.screenr.com/embed/Fps8" width="650" height="396" frameborder="0"></iframe>

# Install Packages
- Are you running RStudio in administrator mode on Windows Vista or Windows 7?
- If not, you need to be. Right click on RStudio and then click "Run as Administrator"
- If you are on a Mac or Windows XP you can disregard this
- Now, copy and paste the code below into the bottom left window in RStudio (the R terminal):
```{r, eval=FALSE,echo=TRUE}
install_new<-function(mypkg){
  if (mypkg %in% installed.packages()) cat("Package already installed")
  else{cat("Package not found, so installing with dependencies... Press CTRL C to abort.")
    Sys.sleep(5)
    install.packages(mypkg,repos="http://cran.wustl.edu/")
}
}

install_new('plyr')
install_new('lmtest')
install_new('ggplot2')
install_new('gridExtra')
install_new('stringr')
install_new('knitr')
install_new('quantreg')
install_new('zoo')

```

# Alternate Install
- If you want or if you have troubles with the above, you can overwrite any existing versions of those packages and do the install in one line, shown below:

```{r eval=FALSE,echo=TRUE}
install.packages(c('plyr','lmtest','ggplot2','gridExtra','stringr'),repos="http://cran.wustl.edu/")
```
```

# Run RStudio
- You should be able to run RStudio now and load any of the packages above.
- Run the code below in the RStudio terminal window (bottom left) to test this:
```{r eval=FALSE,echo=TRUE}
library(lmtest)
example(gqtest)
```

# And you should see...
```{r eval=TRUE,echo=FALSE}
library(lmtest)
example(gqtest)
```

# And test the graphics...
Type in in the terminal (or copy and paste)
```{r echo=TRUE,eval=FALSE}
library(ggplot2)
y<-rt(200,df=5)
qplot(sample=y,stat="qq")
```

# Results
```{r,echo=TRUE,fig.width=9,fig.height=6}
library(ggplot2)
y<-rt(200,df=5)
qplot(sample=y,stat="qq")
```

# That's all
- Did you see the output like the slides above?
- Did you see the plot in the lower right? 

# You are ready to go!
See you on August 2nd and 3rd to find out how to go from these basic steps to using R to learn from your data. Can't wait to see you in Madison!
