% Tutorial 4: Cleaning and Merging Data
% R Bootcamp HTML Slides
% Jared Knowles

# Overview
In this lesson we hope to learn:
- The Strategic Data Project
- Checking data for errors
- Recoding data and changing data types
- Diagnostics and error checks

```{r setup, include=FALSE}
# set global chunk options
opts_chunk$set(fig.path='figure/slides3-', cache.path='cache/slides4-',fig.width=8,fig.height=6,message=FALSE,error=FALSE,warning=FALSE,echo=TRUE)
```

# Data Setup
- Let's read in a new dataset now that has some messiness to it

```{r dataread}
load('data/Student_Attributes.rda')
head(stuatt[,1:4],7)
```
- What's wrong with this?

# How can R help correct this?
- Identify problems
- Enforce business rules for messy data consistently
- Build data cleaning into all analyses tasks across the workflow
- Analyze inconsistencies and do reports

# Strategic Data Project
- The Strategic Data Project is 
- 

# Toolkit--Data Cleaning
- The SDP has come up with a great tutorial and guided analyses using a great synthetic data set to help walk through the process of cleaning data
- This was written in Stata, we have ported it to R, and are going to walk through just a single lesson of it here



