% Tutorial 8: Advanced Topics
% R Bootcamp HTML Slides
% Jared Knowles

# Overview
In this lesson we hope to learn:
- Coding Style
- Git, GitHub, and add-on packages
- Write a `for' loop
- Write a basic function
- Optimize R with parallelization
- Mixed effect models
- Data mining with R

# Basic Principles
- Good professional data analysis is computer programming
- An analysis project is computer code
- Computer code has bugs and to fight bugs the code has to be readable
- 

# Coding Style
- An important part of making work reproducible is making your code readable and understandable
- Much of this involves looking into where to break lines and how to simplify expressions without making them obscure
- You can also take some cues about how to make your code consistent and clean

# For Looping
- Don't do it

# Write a simple function
- Functions are easy
- To view any function in R just type `print(myfunction)`
- For speed, some functions are not viewable

```{r reviewingfunction}
print(mean) #bytecode, we can't see it
print(order)
```


# Git and GitHub
- Working alone or in a group, `git` and GitHub can help
- [Learn Git and GitHub](http://try.github.com/)

# Exercises 
1.
2.

# References


# Session Info

It is good to include the session info, e.g. this document is produced with **knitr** version `r packageVersion('knitr')`. Here is my session info:

```{r session-info}
print(sessionInfo(), locale=FALSE)
```
