% Tutorial 2: Getting Data In
% R Bootcamp HTML Slides
% Jared Knowles

# Overview
- Loading Packages
- Data import is incredibly flexible and once you are used to it, it is easy
- Understand data types
- Load a CSV file
- Organize an analysis project
- Query a database

# A quick note on R `packages`
- `packages` are essentially free and open source add-ons for R
- There are over 3,000 packages available for R that add all sorts of functionality
- A few examples (from the mundane to the crazy)
  * Additional graphics capabilities from the `ggplot2` package
  * Advanced regression techniques from the `lme4` package for mixed effects models
  * 3d graphics from the `scatterplot3d` package (also `webGL`)
  * GIS analytics and mapping functionality with `sp`
  * Text mining analytics with `tm`
  * Predictive modeling frameworks with `caret`
  * Interfaces to other programming languages like Python, Java, and C and C++
  * A web server: `Rserve`
  * And Minesweeper from the `fun` package
  

# I can haz packages?
```{r eval=FALSE,echo=TRUE}
# You can find and install packages within R
install.packages('foo') # Name must be in quotes
install.packages(c('foo','foo1','foo2'))
# Packages get updated FREQUENTLY
update.packages() # Gonna update them all
```
- Note, on Windows Vista and later R either needs to be run as an administrator to install packages, or you have to fiddle with where the packages are installed
- Packages are stored in something called the **library** which is just a collection of packages
  * Sometimes folks call packages **libraries**

# Finding Packages
- Official packages are found on CRAN (Comprehensive R A Network)
- Unofficial packages or **beta** versions of packages are found on RForge and GitHub mostly
- To find out what packages are out there that do a specific function, try:
  * Google "doing X in R package"
  * Look at CRAN taskviews
- CRAN taskviews are great to find a bunch of packages related to a problem you are trying to solve

# Some Must Have Packages
`plyr` `ggplot2` `lme4` `sp` `knitr`

# Data Management
- Data management in R used to be managed by the `ls()` command 
- Go ahead, type it. 
- Now you can look at the Workspace tab in RStudio and have a complete list of the data in R's memory that is accessible to you
- All data objects have names
- All object names are unique (not strictly so, but let's not violate this)
- To reference items within an object type we need to give it an address like in `mydata$thingIwant` or `mydata@thingIwant`
- The `$` and `@` distinction depends on whether this is an S3 or an S4 class

# Ground Rules
- Get used to plain text input files
  * R can handle other formats, but your error rate increases as does the tweaking necessary
- R has a limited set of special characters (symbols) you can use in your data input to be translated correctly 
- These symbols are reserved and will be interpreted in strange ways if you include them in your plain text data file
- Most of them are fairly obvious operators, see [Paul Murrell's excellent summary](http://www.stat.auckland.ac.nz/~paul/ItDT/HTML/node75.html)


# Missing Data Symbols
- Missing data has the symbols `NA` or `NaN` or `NULL` depending on the context.
- Consider:
```{r missing}
a<-c(1,2,3)  # a is a vector with three elements
# Ask R for element 4
print(a[4])
```
- But what is the difference between `NA` and `NULL`?
```{r nulldata}
ls() # get objects
a<-c(a,NULL) # Append NULL onto a
print(a)
# Notice no change
a<-c(a,NA)
print(a)
```
- `NA` can hold a place, `NULL` cannot
- `NaN` is even more special, and only holds things like imaginary numbers
```{r nan, warning=TRUE}
b<-1
b<-sqrt(-b)
print(b)
pi/0
```
- Inf is a special case as well representing an infinite value; just for fun `sin(Inf)` = `NaN`

# Beginning Analysis
- Now let's set up our analysis project
- It is best to keep projects discrete in directories
- Create a few subdirectories
  - Data
  - Functions / R src
  - Figures / Plots
  - Cleaned Data
- See the ProjectTemplate package for a more detailed philosophy about organizing projects

# Organization of a project is key
- Separate data from scripts
- Separate automatic scripts from interactive scripts
- Put figures apart from both of these
- **Always keep your raw data**

# Read in Data
- Reading in data is one of the trickiest issues for R
- This is because R is incredibly flexible and can handle data in almost any form including `.csv` `.dta` `.sas` `.spss` `.dat` and even `.xls` and `.xlsx` with some care
- So we have to carefully specify the data types to R so it can understand what form the data needs to take
- Compared to C this is great!

# CSV is Our Friend
- The easiest data type is .csv though Excel files can be read as well `r load('data/smalldata.rda')`
```{r readdata, eval=FALSE}
# Set working directory to the tutorial directory
# In RStudio can do this in "Tools" tab
setwd('~/GitHub/r_tutorial_ed')
#Load some data
df<-read.csv('data/smalldata.csv')
# Note if we don't assign data to 'df'
# R just prints contents of table
```

# Let's Check What We Got
```{r checkresults,echo=FALSE}
# Let's see what object types 
# R assigned to our dataset
str(df[27:32])
```

# Always Check Your Data
- A few great commands:
```{r dim}
  dim(df)
```
-
```{r summary}
  summary(df[,1:5])
```
-
```{r names}
  names(df)
```
-
```{r attributes}
  names(attributes(df))
  class(df)
```
  

# Data Warehouses, Oracle, SQL and RODBC
- Do you have data in a warehouse? 
  - RODBC can help
- You can query the data directly and bring it into R, saving time and hassle
- Makes your work reproducible, always start with a clean slate of data
- At DPI this can allow us to pull data directly from LDS or other databases using SQL queries

# An Example From DPI
- The basics of the RODBC package are easy to understand

```{r rodbc, eval=FALSE}
library(RODBC) # interface driver for R
channel<-odbcConnect("Mydatabase.location",uid="useR",pwd="secret") # establish connection
# we can do multiple connections in the same R session
# credentials stored in plain text unless you do some magic
table_list<-sqltables(channel,schema="My_DB") # Get a list of tables in the connection
colnames(sqlFetch(channel,'My_DB.TABLE_NAME',max=1)) # get the column names of a table

datapull<-sqlQuery(channel,"SELECT DATA1, DATA2, DATA3 FROM My_DB.TABLE_NAME") # execute some SQLquery, can paste any SQLquery as a string into this space
```

# Missing Data
- Let's add some missing data to our dataframe so we can see how missing data works

```{r munging}
random<-sample(unique(df$stuid),100)
random2<-sample(unique(df$stuid),120)
messdf<-df
messdf$readSS[messdf$stuid %in% random]<-NA
messdf$mathSS[messdf$stuid %in% random2]<-NA
```

# Checking for Missing Data
- The `summary` function helps identify missing data
```{r summarym}
summary(messdf[,c('stuid','readSS','mathSS')])
nrow(messdf[!complete.cases(messdf),]) # number of rows with missing data
```
- To get rid of missing data, we can copy our data with all missing cases dropped using the `na.omit` function
```{r naomit}
cleandf<-na.omit(messdf)
nrow(cleandf)
```

# Now we have the data
- What next?
- We need to do some basic diagnostics on our data to understand the look and feel of it before we proceed
- Here are a few examples of scripts we could run to understand our data object
```{r datafirstcut}
dim(messdf)
str(messdf[,18:26])
names(messdf)
```
- It looks like we have a number of `id` variables, this is useful and it is good to check if these variables have multiple rows per id or not and we do this using `length` and `unique`
```{r checkunique}
length(unique(messdf$stuid))
length(unique(messdf$schid))
length(unique(messdf$dist))
```

# Checking for Coding
- Data is coded using numeric or character representations of attributes--commonly things are coded using a 1 and 0 scheme or an A,B,C scheme
- With R we can check how our variables are coded very easily

```{r checkcoding}
unique(messdf$grade)
unique(messdf$econ)
unique(messdf$race)
unique(messdf$disab)
```

# Next Steps
- In the next section we will learn to aggregate, explore, reshape, and recode data
- Questions?

# Exercises
1.
2.
3. 

# Other References
- [UCLA Academic Technology Services: Reading in Raw Data](http://www.ats.ucla.edu/stat/r/pages/raw_data.htm)
- [An R Vocabulary for Starting Out](https://github.com/hadley/devtools/wiki/vocabulary)
- [Quick-R: Data Import](http://www.statmethods.net/input/importingdata.html)
- [Video Tutorials](http://www.twotorials.com/)


# Session Info

It is good to include the session info, e.g. this document is produced with **knitr** version `r packageVersion('knitr')`. Here is my session info:

```{r session-info}
print(sessionInfo(), locale=FALSE)
```