% R Bootcamp HTML Slides
% Jared Knowles
% 05/13/2012

```{r setup, include=FALSE}
# set global chunk options
opts_chunk$set(fig.path='figure/slides-', cache.path='cache/slides-', cache=TRUE)
# upload images automatically
opts_knit$set(upload.fun = imgur_upload)
```

# R
- R is an Open Source (and freely available) environment for statistical computinga nd graphics
- Available for Windows, Mac OS X, and Linux
- R is being actively developed with two major releases per year and dozens of releases of add on packages
- R can be extended with 'packages' that contain data, code, and documentation to add new functionality

# Why Use R
- R is a common tool among data experts at major universities
- No need to go through procurement, R can be installed in any environment on any machine and used with no licensing or agreements needed
- R source code is very readable to increase transparency of processes
- R code is easily borrowed from and shared with others
- R is incredibly flexible and can be adapted to specific local needs
- R is under incredibly active development, improving greatly, and supported wildly by both professional and academic developers

# R Advantages Continued
- R is platform agnostic--Linux, Mac, PC, server, desktop, etc.
- R can output results in a variety of formats
- R can build routines straight out of a database for common and universal reporting

R Can Compliment Other Tools
------------------------------------
- R plays nicely with data from Stata, SPSS, SAS and others
- R can check work, produce output, visualize results from other programs
- R can do bleeding edge analysis that aren't available in proprietary packages yet
- R is becoming more prevalent in undergraduate statistics courses

# R As A Calculator

```{r computing,echo=TRUE,results='markup'}
2+2 # add numbers
2*pi #multiply by a constant
7+runif(1,min=0,max=1) #add a random variable
4^4 # powers
sqrt(4^4) # functions
```

# Using the Workspace
- To do more we need to learn how to manipulate the 'workspace'.
- This includes all the scalars, vectors, datasets, and functions stored in memory.
- All R objects are stored in the memory of the computer, limiting the available space for calculation to the size of the RAM on your machine.
- R makes organizing the workspace easy.
```{r}
x<-5 #store a variable with <-
x    #print the variable
z<-3 
ls() #list all variables
ls.str() #list and describe variables
rm(x)    # delete a variable
ls()
```

# R as a Language
- R is more than statistical software, it is a computer language
- Like any language it has rules (some poorly enforced), and conventions
- You will learn more as you go, but we'll go over a few to start

1. Case sensitivity matters
```{r}
a<-3
A<-4
print(c(a,A))
```
  * <font color="red">**a** &#8800; **A**</font> 
2. What happens if I type **print(a,A)**?
  * So what does **c** do?
```{r}
A<-c(3,4)
print(A)
```
  * **c** stands for concatenate and allows vectors to have multiple elements

# Vectors 
- Everything is a vector in R, even single numbers
```{r vectors}
print(1)
# The 1 in braces means this element is a vector of length 1
print("This tutorial is awesome")
# This is a vector of length 1 consisting of a single "string of characters"
print(LETTERS) 
# This vector has 26 character elements
print(LETTERS[6])
# The sixth element of this vector has length 1
length(LETTERS[6])
# The length of that element is a number with length 1
```

# Language
- In language there are a number of ways to say the same thing
  *  <font color="green">The dog chased the cat.</font> 
  *  <font color ="blue">The cat was chased by the dog.</font>
  *  <font color ="red">By the dog, the cat was chased.</font>
- Some ways are more elegant than others, all convey the same message. 

```{r language}
a<-runif(100) # Generate 100 random numbers
b<-runif(100) # 100 more
c<-NULL       # Setup for loop (declare variables)
for(i in 1:100){  # Loop just like in Java or C
c[i]<-a[i]*b[i]
}
d<-a*b
identical(c,d) # Test equality
```
- Which is nicer?

# Reading Data In
- To read data in we have to tell R where it currently is on the filesystem by setting a working directory
- Then we have to tell it where to look for the dataset and what format that dataset is in
- CSV files are **simplest** for beginning use cases, but R is flexible
```{r}
# Set working directory to the tutorial director
# In RStudio can do this in "Tools" tab
setwd('~/r_tutorial_ed')
#Load some data
df<-read.csv('data/smalldata.csv')
# Note if we don't assign data to 'df'
# R just prints contents of table
```

# Objects
- Everything in R is an object--even functions
- Objects can be manipulated many ways
- A common example is applying the `summary' function to a variety of object types and seeing how it adapts
```{r}
summary(df[,28:31]) #summary look at df object
summary(df$readSS) #summary of a single column
```
-The **$** says to look for object **readSS** in object **df**

# Graphics too

```{r graphics, message=FALSE, fig.cap='Student Test Scores', fig.width=7, fig.height=5}
library(ggplot2) # Load graphics Package
qplot(readSS,mathSS,data=df,geom='hex')+theme_bw()+opts(title='Test Score Relationship')
```

# Handling Data in R
- R handles data differently than many other statistical packages
- In R, all elements are objects
```{r}
length(unique(df$school))
length(unique(df$stuid))
```
- Results of function calls can be stored


# How
- source editor: [RStudio](http://www.rstudio.org/) (perfect integration with [**knitr**](http://yihui.name/knitr/); one-click compilation); currently you have to use the [preview version](http://www.rstudio.org/download/preview) (>= 0.96.109)
- HTML5 slides converter: [pandoc](http://johnmacfarlane.net/pandoc/); this document was generated by: `pandoc -s -S -i -t dzslides --mathjax knitr-slides.md -o knitr-slides.html`
- the file [`knitr-slides.md`](https://github.com/yihui/knitr/blob/master/inst/examples/knitr-slides.md) is the markdown output from its [source](https://github.com/yihui/knitr/blob/master/inst/examples/knitr-slides.Rmd): `library(knitr); knitr('knitr-slides.Rmd')`
- or simple click the button `Knit HTML` in RStudio

# For ninjas

- you should tweak the default style; why not try some [Google web fonts](http://www.google.com/webfonts)? (think how painful it is to wrestle with fonts in LaTeX)
- pandoc provides 3 types of HTML5 slides (dzslides is one of them)
- you can tweak the default template to get better appearances


# Reproducible research

It is good to include the session info, e.g. this document is produced with **knitr** version `r packageVersion('knitr')`. Here is my session info:

```{r session-info}
print(sessionInfo(), locale=FALSE)
```