% Tutorial 2: Getting Data In
% R Bootcamp HTML Slides
% Jared Knowles

# Overview
Getting data in is the highest hurdle in R
- Data import is incredibly flexible and once you are used to it, it is easy
- Load a CSV file
- Organize an analysis project
- Query a database

# Ground Rules
- Get used to plain text input files
  * R can handle other formats, but your error rate increases as does the tweaking necessary
- R has a limited set of special characters (symbols) you can use in your data input to be translated correctly 
- These symbols are reserved and will be interpreted in strange ways if you include them in your plain text
- Most of them are fairly obvious operators, see [Paul Murrell's excellent summary](http://www.stat.auckland.ac.nz/~paul/ItDT/HTML/node75.html)


# Missing Data Symbols
- Missing data has the symbols `NA` or `NaN` or `NULL` depending on the context.
- Consider:
```{r missing}
a<-c(1,2,3)  # a is a vector with three elements
# Ask R for element 4
print(a[4])
```
- But what is the difference between `NA` and `NULL`?
```{r nulldata}
ls() # get objects
a<-c(a,NULL) # Append NULL onto a
print(a)
# Notic eno change
a<-c(a,NA)
print(a)
# NA can hold a place, NULL cannot
```
- `NaN` is even more special, and only holds things like imaginary numbers
```{r nan}
b<-1
b<-sqrt(-b)
print(b)
```


