% Tutorial 3: Manipulating Data in R
% R Bootcamp HTML Slides
% Jared Knowles

# Overview
In this lesson we hope to learn:
- Aggregating data
- Organizing our data
- Manipulating vectors
- Dealing with missing data


```{r setup, include=FALSE}
# set global chunk options
opts_chunk$set(fig.path='figure/slides3-', cache.path='cache/slides3-',fig.width=8,fig.height=6,message=FALSE,error=FALSE,warning=FALSE,echo=TRUE,dev='svg',size='tiny')
```


# Again, read in our dataset
```{r readdata}
# Set working directory to the tutorial directory
# In RStudio can do this in "Tools" tab
setwd('~/GitHub/r_tutorial_ed')
#Load some data
load("data/smalldata.rda")
df<-mydat
rm(mydat)
# Note if we don't assign data to 'df'
# R just prints contents of table
```

# Aggregation
- Sometimes we need to do some basic checking for the number of observations or types of observations in our dataset
- To do this quickly and easily--the `table` function is our friend
- Let's look at our observations by year and grade
```{r table}
table(df$grade,df$year)
```
- The first command gives the rows, the second gives the columns

# Aggregation can be more complex
- Let's aggregate by race and year

```{r table2}
table(df$year,df$race)
```

- Race is consistent across years, interesting
- What if we want to only look at 3rd graders that year?

# More complicated still

```{r table3}
with(df[df$grade==3,],{table(year,race)})
```
- `with` specifies a data object to work on, in this case all elements of `df` where `grade==3`
- `table` is the same command as above, but since we specified the data object in the `with` statement, we don't need the `df$` in front of the variables of interest

# Quick exercise
- Can you find the number of black students in each grade in each year?
- hint: `with(df[df$___=="B",]...)`
- How many in year 2002, grade 6?
  * 48
- How many in 2001, grade 7?
  * 39


# Answer
```{r table4}
with(df[df$race=="B",],{table(year,grade)})
```

# Tables cont.
- This is really powerful for looking at the descriptive dimensions of the data, we can ask questions like:
- how many students are at each proficiency level each year?
```{r proftable}
table(df$year,df$proflvl)
```
- how many students are at each proficiency level by race?
```{r raceproftable}
table(df$race,df$proflvl)
```

# Proportional Tables
- What if we aren't interested in counts? 
- R makes it really easy to calculate proportions
```{r proptable1}
prop.table(table(df$race,df$proflvl))
```
- Hmmm, this is goofy. This tells us the proportion of each cell out of the total. Also, the digits are distracting. How can we fix this?

# Try number 2
```{r proptable2}
round(prop.table(table(df$race,df$proflvl),1),digits=3)
```
- The `1` tells R we want proportions rowise, a `2` goes columnwise
- A few more problems arise--this pools all observations, including students across years

# Aggregating Data
- One of the most common questions will be to compute aggregates of data
- R has an `aggregate` function that can be used and helps us avoid the clustering problems above
- This works great for simple aggregation like scale score by race, we just need a `formula` (think I want variable X **by** grouping factor Y) and the statistic we want to compute

```{r aggregate1}
# Reading Scores by Race
aggregate(readSS~race,FUN=mean,data=df)
```

# Aggregate (II)
- `aggregate` can take us a little further, we can use aggregate multiple variables at a time
```{r aggregate2}
aggregate(cbind(readSS,mathSS)~race,data=df,mean)
```
- We can add multiple grouping varialbes using the `formula` syntax
```{r aggregate3}
head(aggregate(cbind(readSS,mathSS)~race+grade,data=df,mean),8)
```

# Crosstabs
- We can build a systematic cross-tab now
```{r crosstab}
ag<-aggregate(readSS~race+grade,data=df,mean)
xtabs(readSS~., data=ag)
```
- And prettier output
```{r crosstabpretty}
ftable(xtabs(readSS~.,data=ag))
```

# Check your work
- What is the mean reading score for 6th grade students with disabilities?
  * __481.83__
- How many points is this from non-disabled students?
  * __29.877__


# Answer II
```{r aggregatetest}
aggregate(cbind(readSS,mathSS)~disab+grade,data=df,mean)
```

# Aggregate Isn't Enough
- `aggregate` is cool, but it isn't very flexible
- We can only use aggregate output as a table, which we have to convert to a data frame
- There is a better way; the `plyr` package
- `plyr` is a set of routines/logical structure for transforming, summarizing, reshaping, and reorganizing data objects of one type in R into another type
- We will focus here on summarizing and aggregating a data frame, but later in the bootcamp we'll apply functions to lists and turn lists into data frames as well
- This is cool!

# School Means
- Consider the case we want to turn our student level data into school level data
- Who hasn't had to do this?!?
- In `aggregate` we do:
```{r aggregate4}
z<-aggregate(readSS~dist,FUN=mean,data=df)
z
```
- But I want more! I want to aggregate multiple variables. I want to do it across multiple groups. I want the output to be a dataframe I can work on.
- Thank you `plyr`

# Using plyr
- `plyr` has a straightforward syntax
- All `plyr` functions are in the format **XX**ply. The two X's specify what the input file we are applying a function to is, and then what way we would like it outputted.
- In `plyr` d = dataframe, l= list, m=matrix, and a=array. By far the most common usage is `ddply`
- From a dataframe, to a dataframe.

# plyr in Action
```{r plyr1}
library(plyr)
myag<-ddply(df, .(dist,grade),summarize,
            mean_read=mean(readSS,na.rm=T),
            mean_math=mean(mathSS,na.rm=T),
            sd_read=sd(readSS,na.rm=T),
            sd_math=sd(mathSS,na.rm=T),
            count_read=length(readSS),
            count_math=length(mathSS))
head(myag)
```

# More plyr
- This is great, we can quickly build a summary dataset from individual records
- A few advanced tricks. How do we build counts and percentages into our dataset?
```{r plyr3}
myag<-ddply(df, .(dist,grade),summarize,
            mean_read=mean(readSS,na.rm=T),
            mean_math=mean(mathSS,na.rm=T),
            sd_read=sd(readSS,na.rm=T),
            sd_math=sd(mathSS,na.rm=T),
            count_read=length(readSS),
            count_math=length(mathSS),
            count_black=length(race[race=='B']),
            per_black=length(race[race=='B'])/length(readSS))
summary(myag[,7:10])            
```

# Quick Check
- What if we want to compare how districts do on educating ELL students?
- What district ID has the highest mean score for 4th grade ELL students on reading? Math?
  * 66 in reading, 105 in math
- How many students are in these classes?
  * 12 and 7 respectively


# Answer III
```{r check2}
myag2<-ddply(df, .(dist,grade,ell),summarize,
            mean_read=mean(readSS,na.rm=T),
            mean_math=mean(mathSS,na.rm=T),
            sd_read=sd(readSS,na.rm=T),
            sd_math=sd(mathSS,na.rm=T),
            count_read=length(readSS),
            count_math=length(mathSS),
            count_black=length(race[race=='B']),
            per_black=length(race[race=='B'])/length(readSS))
subset(myag2,ell==1&grade==4)        
```

# Sorting
- A key way to explore data in tabular form is to sort data
- Sorting data in R can be dangerous as you can reorder the vectors of a dataframe
- We use the `order` function to sort data

```{r sortwrong}
df.badsort<-order(df$readSS,df$mathSS)
head(df.badsort)
```
- Why is this wrong?
- What is R giving us?
- Rownames...

# Correct Example
- To fix it, we need to tell R to reorder the rownames in the order we want

```{r sortright}
df.sort<-df[ order(df$readSS,df$mathSS,df$attday),]
head(df[,c(3,23,29,30)])
head(df.sort[,c(3,23,29,30)])
```

# Let's clean it up a bit more
```{r sortright2}
head(df[with(df,order(-readSS,-attday)),c(3,23,29,30)])
```
- Here we find the high performing students, note that the `-` denotes we want descending order, R's default is ascending order
- This is easy to correct

# About sorting
- Sorting works differently on some data types, matrices are slightly different
```{r matrixsort}
M<-matrix(c(1,2,2,2,3,6,4,5),4,2,byrow=FALSE,dimnames=list(NULL,c("a","b")))
M[order(M[,"a"],-M[,"b"]),] # can't use "with"
```
- Tables are familiar
```{r sorttable}
mytab<-table(df$grade,df$year)
mytab[order(mytab[,1]),]
mytab[order(mytab[,2]),]
```

# Filtering Data
- Filtering data is an incredibly powerful feature and we have already seen it used to do some interesting things
- Filtering data in R is loaded with trouble though, because the filtering arguments must be very carefully specified
- Filtering is like a mini-sort
- Always, always, always check your work
- And remember, this is the place the NAs do the most damage
- Let's look at some examples

# Basic Filtering a Column
```{r columnfilter}
# Gives all rows that meet this requirement
df[df$readSS>800,] 
df$grade[df$mathSS>800] 
# Gives all values of grade that meet this requirement
```
- This seems basic enough, let's filter on multiple dimensions
- Before the brackets we specify what we want returned, and within the brackets we present the logical expression to evaluate
- Behind the scenes R does a logical test and gets the row numbers that match the logical expression
- It then combines them back with the object in front of the brackets to return the values
- So great we don't have to do that!

# Multiple filters
```{r filterlayers}
df$grade[df$black==1 & df$readSS>650]
```
- What happens if we type `df$black=1` or `black==1`? 
- Why won't this work?

# Using filters to assign values
- We can also use filters to assign values as well
- This is how you recode variables and create new ones
- Let's create a variable `spread` indicating whether a district has high or low spread among its student scores
```{r filtersort}
myag$spread<-NA # create variable
myag$spread[myag$sd_read<75]<-"low"
myag$spread[myag$sd_read>75]<-"high"
myag$spread<-as.factor(myag$spread)
summary(myag$spread)
```

# Check your work
- The previous block of code is a useful way to learn how to recode variables

```{r recodeexamp,eval=FALSE}
myag$spread<-NA # create variable
myag$spread[myag$sd_read<75]<-"low"
myag$spread[myag$sd_read>75]<-"high"
myag$spread<-as.factor(myag$spread)
```

- Create a new variable in `myag` called `schoolperf` for `mean_math` scores with the following coding scheme:


Grade | Score Range | Code
----- | ----------- | ----
3     |   >425      | "Hi"
4     |   >450      | "Hi"
5     |   >475      | "Hi"
6     |   >500      | "Hi"
7     |   >525      | "Hi"
8     |   >575      | "Hi"

- All other values are coded as "lo"
- How many "hi" and "lo" observations do we have?
- By `dist`?

# Results
```{r answercoding}
myag$schoolperf<-"lo"
myag$schoolperf[myag$grade==3 & myag$mean_math>425]<-"hi"
myag$schoolperf[myag$grade==4 & myag$mean_math>450]<-"hi"
myag$schoolperf[myag$grade==5 & myag$mean_math>475]<-"hi"
myag$schoolperf[myag$grade==6 & myag$mean_math>500]<-"hi"
myag$schoolperf[myag$grade==7 & myag$mean_math>525]<-"hi"
myag$schoolperf[myag$grade==8 & myag$mean_math>575]<-"hi"
myag$schoolperf<-as.factor(myag$schoolperf)
summary(myag$schoolperf)
table(myag$dist,myag$schoolperf)
```


# Let's replace data
- For district 6 let's negate the grade 3 scores by replacing them with missing data
```{r replacedata}
myag$mean_read[myag$dist==6 &myag$grade==3]<-NA
head(myag[,1:4],2)
```
- Let's replace one data element with another
```{r replacedata2}
myag$mean_read[myag$dist==6 & myag$grade==3]<-myag$mean_read[myag$dist==6 & myag$grade==4]
head(myag[,1:4],2)
```
- Voila

# Why do NAs matter so much?
- Let's consider the case above but insert some NA values for all 3rd grade tests
```{r munge}
myag$mean_read[myag$grade==3]<-NA
head(myag[order(myag$grade),1:4])
```


# NAs II
- Now let's calculate a few statistics:
```{r means}
mean(myag$mean_math)
mean(myag$mean_read)
```
- Remember, NA values propogate, so R assumes an NA value could take literally any value, and as such it is impossible to know the `mean` of a vector with NA
- We can override this though:
```{r meansna}
mean(myag$mean_math,na.rm=T)
mean(myag$mean_read,na.rm=T)
```

# Beyond the Mean
- But for other problems it is tricky
- What if we want to know the number of rows that have a `mean_read` of less than 500?
```{r length}
length(myag$dist[myag$mean_read<500])
head(myag$mean_read[myag$mean_read<500])
```
- And what if we want to add the standard deviation to these vectors?
```{r addtomissing}
badvar<-myag$mean_read+myag$sd_read
summary(badvar)
```

# So we need to filter NAs explicitly
- Consider the case where two sets of variables have different missing elements
```{r moremissing}
myag$sd_read[myag$count_read<100 & myag$mean_read<550]<-NA
length(myag$mean_read[myag$mean_read<550])
length(myag$mean_read[myag$mean_read<550 & !is.na(myag$mean_read)])
```
- What is `!is.na()` ?
  * `is.na()` is a helpful function to identify TRUE if a value is missing
  * `!` is the reverse operator
  * We are asking R if this value is not a missing value, and to only give us non-missing values back

# Merging Data
- It is unlikely all the data we will want resides in a single dataset and often we have to combine data from several sources
- R makes this easy, but that simplicity comes at a cost--it can be easy to make mistakes if you don't specify things carefully

```{r createwidedata,echo=FALSE,results='hide'}
widedf<-reshape(df,timevar="year",idvar="stuid",direction="wide")
```


# Reshaping Data
- Reshaping data is a slightly different issue than aggregating data
- Let's review the two data types: long and wide
```{r longpreview}
head(df[,1:10],3)
```
- Now let's look at wide:
```{r widepreview}
head(widedf[,28:40],3)
```
- How did we do this?

# Wide data v. Long Data 
- The great debate
- Most econometrics, panel, and time series datasets come wide and so these seem familiar
- R for most cases prefers long data, including for most graphing and analysis functions
- But, not all
- So we have to learn both

# The reshape Function
- `reshape` is the way to move from wide to long
- The data stays the same, but the shape of it changes
- The long data had dimensions: `r dim(df)`
- The wide data has dimensions: `r dim(widedf)`
- How do we get to these numbers?
  * The rows in the wide dataframe represent unique students

# Deconstructing reshape
```{r reshapeanaly, eval=FALSE}
widedf<-reshape(df,timevar="year",idvar="stuid",direction="wide")
```
- `idvar` represents the unit we want to represent a single row, in this case each unique student gets a single row
- In this simple case `timevar` is the variable that differenaties between two rows with the same student ID
- `direction` tells R we are going to move to wide data
- As written all data will move, but using the `varying` argument we can tell R explicitly which items we want to move wide

# What about Wide to Long?
- We often need to do this to plot data in R
- Luckily the `reshape` function works well in both directions
```{r widetolong}
longdf<-reshape(widedf,idvar='stuid',timevar='year',
varying=names(widedf[,2:91]),direction='long',sep=".")
```
- If our data is formatted nicely, R can do the guessing and identify the years for us by parsing the dataframe names


# Subsetting Data
- We have already seen a lot of subsetting examples above, which is what filtering is, but R provides some great shortcuts to this
- Let's look at the `subset` function to get only 4th grade scores
```{r subset1}
g4<-subset(df,grade==4)
dim(g4)
```
- This is equivalent to:
```{r subset2}
g4_b<-df[df$grade==4,]
```
- These two elements are the same:
```{r testofequality}
identical(g4,g4_b)
```

# That's it
- Now you can filter, subset, sort, recode, and aggregate data!
- Let's look at a few exercises to test these skills
- Once these skills are mastered, we can begin to understand how to automate R to clean data with known errors, and to recode data in R so it is ready to be used for analysis
- Then we can really take off!

# Exercises
1. Sort `df` on `measerr` and `mathss`. What are the highest 5 values of each. 



# Other References
- [An R Vocabulary for Starting Out](https://github.com/hadley/devtools/wiki/vocabulary)
- [Quick-R: Data Management](http://www.statmethods.net/management/index.html)
- [UCLA ATS: R FAQ on Data Management](http://www.ats.ucla.edu/stat/r/faq/default.htm)
- [Video Tutorials](http://www.twotorials.com/)

