% Tutorial 2: Getting Data In
% R Bootcamp HTML Slides
% Jared Knowles

# Overview
Getting data in is the highest hurdle in R
- Data import is incredibly flexible and once you are used to it, it is easy
- Understand data types
- Load a CSV file
- Organize an analysis project
- Query a database

# Ground Rules
- Get used to plain text input files
  * R can handle other formats, but your error rate increases as does the tweaking necessary
- R has a limited set of special characters (symbols) you can use in your data input to be translated correctly 
- These symbols are reserved and will be interpreted in strange ways if you include them in your plain text
- Most of them are fairly obvious operators, see [Paul Murrell's excellent summary](http://www.stat.auckland.ac.nz/~paul/ItDT/HTML/node75.html)


# Missing Data Symbols
- Missing data has the symbols `NA` or `NaN` or `NULL` depending on the context.
- Consider:
```{r missing}
a<-c(1,2,3)  # a is a vector with three elements
# Ask R for element 4
print(a[4])
```
- But what is the difference between `NA` and `NULL`?
```{r nulldata}
ls() # get objects
a<-c(a,NULL) # Append NULL onto a
print(a)
# Notic eno change
a<-c(a,NA)
print(a)
# NA can hold a place, NULL cannot
```
- `NaN` is even more special, and only holds things like imaginary numbers
```{r nan}
b<-1
b<-sqrt(-b)
print(b)
```

# Beginning Analysis
- Now let's set up our analysis project
- It is best to keep projects discrete in directories
- Create a few subdirectories
  - Data
  - Functions / R src
  - Figures / Plots
  - Cleaned Data
- See the ProjectTemplate package for a more detailed philosophy about organizing projects

# Organization of a project is key
- Separate data from scripts
- Separate automatic scripts from interactive scripts
- Put figures apart from both of these
- **Always keep your raw data**

# Read in Data
- Reading in data is one of the trickiest issues for R
- This is because R is incredibly flexible and can handle data in almost any form
- So we have to carefully specify the data types to R so it can understand what form the data needs to take
- Compared to C this is great!

# CSV is Our Friend
- The easiest data type is .csv though Excel files can be read as well
```{r}
# Set working directory to the tutorial director
# In RStudio can do this in "Tools" tab
setwd('~/GitHub/r_tutorial_ed')
#Load some data
df<-read.csv('data/smalldata.csv')
# Note if we don't assign data to 'df'
# R just prints contents of table
```

# Let's Check What We Got
```{r checkresults,echo=FALSE}
# Let's see what object types 
# R assigned to our dataset
str(df[22:32])
```

# Always Check Your Data
- A few great commands:
  - ```{r dim}
  dim(df)
  ```
  - ```{r summary}
  summary(df[,1:5])
    ```
  -```{r attributes}
  names(attributes(df))
```
  

# Some Slide on RODBC
- Do you have data in a warehouse? 
  - RODBC can help
- You can query the data directly and bring it into R, saving time and hassle
- Makes your work reproducible, always start with a clean slate of data
- At DPI this can allow us to pull data directly from LDS or other databases

# Now we have the data
- What next?